Demographic consequences of partner diversity and turnover in a multi-species ant-plant mutualism

### 2024-03-05 Manuscript/ant_cactus_AmNat.tex
ipm methods

Multiple insect herbivores and seed predators specialize on tree cholla \citep{Mann1969}. 
The Cerambycid beetle \textit{Moneilema appressum} and a weevil (Curculionidae) of the genus \textit{Gerstaekeria} feed on vegetative and reproductive structures as adults and their larvae feed internally. 
A cactus bug, \textit{Narnia pallidicornis} (Hemiptera: Coreidae), feeds on all cholla parts with a preference for the reproductive structures \citep{Miller2006}.
A seed predator, \textit{Cahela ponderosella} (Lepidoptera: Pyralidae), oviposits in open flowers and larvae eat seeds in developing fruits. 
These predators can have significant negative impacts on plant fitness of and depress population growth \citep{Miller2009}.
Prior experiments showed that ant-tended tree cholla experience less herbivory and seed predation than plants from which ants were excluded \citep{Miller2007,Ohm2014}.

\subsection*{Data Collection}
This study is based on long-term demographic data spanning 2004 to 2023 at SEV-LTER. 
From 2004 to 2008, we censused 134 plants distributed across three spatial blocks. 
This initial census group was discontinued in 2009, when we established six 30 $\times$ 30-meter plots and tagged all tree cholla within those plots. 
Two additional 30 $\times$ 30-meter plots were added in 2011, and this group of eight plots has since been censused annually through 2023 (with the exception of 2020 due to the pandemic shutdown). 
For all plants, in May or early June of each year we recorded plant survival since the last survey and, for survivors, we recorded height (cm), maximum crown width (cm), and crown width perpendicular to the maximum (cm).
Size measurements were used to calculate plant volume ($cm^3$) based on the volume of an elliptical cone. 
We recorded reproductive effort as counts of viable and aborted flowerbuds. 
We recorded the ant species present (or vacancy if no ants present).
Occurrences of more than one ant species on one plant were rare (less than 5\% of observations), and for the purpose of this analysis we classified the plant as being occupied by the more abundant species. 
Plots were searched for new recruits each year, and these were added to the census.
In total, the data set included 1141 unique individuals and \tom{19 plant-year observations}{It is 19 years but the number of plant-year observations should be the total number of rows in the data set}. 
%These data were used to fit vital rate models (survival, growth, reproduction) as functions of plant size and ant occupancy state.

We used additional, smaller data sets from previously published studies to estimate seed and seed bank parameters. 
Ohm et al. \citep{Ohm2014} provide data on the number of seeds per fruit for plants tended by \textit{L. apiculatum}, \textit{C. opuntiae}, or no ants (experimental exclusion). 
Miller et al. \citep{Miller2009} provide data on seed entry to the seed bank and seedling germination and survival rates.

\subsection*{Multi-state Integral Projection Model}
Integral Projection Models describe population dynamics in discrete time, with functions that relate vital rates to continuous state variables. 
While IPMs are a natural choice for populations with continuous size structure, they can also be modified to accommodate a combination of continuous and discrete state variables, as we do here. 
We constructed a multi-state IPM that stitches together population structure associated with plant size and ant state, allowing us to determine the individual fitness effects of each ant species and the composite effects of multiple partners, with their transition dynamics modeled explictly.

Given the low frequency of ant species other than \textit{L. apiculatum} and \textit{C. opuntiae} (7.78\% of observed ants) we combined observations of all other ants into an ``other'' category, such that our models included four possible ant states: vacant, \textit{L. apiculatum}, \textit{C. opuntiae}, and ``other''. 
The ``other'' category was made of unidentified ant species (2.8\% of observed ants), unknown species belonging to the geni \textit{Aphenogaster} (0.4\%),  \textit{Camponotus} (0.9\%),  \textit{Tetramorium} (0.02\%), \textit{Brachymyrmer} (0.02\%), a honeypot ant (0.08\%), and \textit{Forelius pruinosus} (3.5\%).
Ant state is included as a predictor variable in sub-models where there are biologically realistic pathways through which ants could impact the outcome of that process. 
For example, ant partners defend cacti from herbivores, and prior experimental work indicates that ant tending can reduce vegetative tissue loss and floral abortion.
Therefore, ant state was included in sub-models for survival, growth, and flowerbud viability. 
In contrast, we have no reason to expect that ant tending can directly influence the probability of flowering and flowerbud production, independently of its influence on plant size. 
Therefore, these sub-models included plant size but not ant state as predictor variables.

Following previous studies, we modeled the tree cholla life cycle using continuously size-structured plants where $n(x,a)_{t}$ gives the number of plants of size $x$  and ant state $a$ in year $t$, plus two discrete seed banks ($B^1_{t}$ and $B^2_{t}$) corresponding to 1 and 2-year old seeds. 
Seed bank dynamics are given by:

\begin{linenomath*}
	$$
	B^1_{t+1} = \delta \sum_{a=1}^{A} \int_L^U  \kappa(a') P(x;\pmb{\tau^P}) F(x;\pmb{\tau^F}) V(a;\pmb{\tau^V_{a}}) n(x,a)_{t} dx \\
	$$
	$$
	B^2_{t+1} =  (1 - \gamma_1)B^1_{t}\\
	$$
\end{linenomath*}

\noindent In these equations $x$ and $x'$ indicate the size of a plant in years $t$ and $t+1$ respectively, $a$ and $a'$ indicate the ant partner of a plant in years $t$ and $t+1$.
Functions $P(x;\pmb{\tau^P})$ and $F(x;\pmb{\tau^F})$ give the probability of flowering in year $t$ and the number of flowerbuds produced in year $t$, respectively, by plants of size $x$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau^V_{a}})$) and the number of seeds per fruit ($\kappa(a)$) are dependent on ant state $a$ but not size. 
The vector $\pmb{\tau}$ gives year-specific deviates (with mean zero) and appears in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and subscripts indicate that deviates are specific to plants in ant state $a$ and year.
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible ant states ($A=4$) giving total seed production. 
Seeds are multiplied by the probability of seed dispersal and survival ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

For the above-ground part of the life cycle, the number of plants of size $x'$ and ant state $a'$ in year $t+1$ ($n(x',a')_{t+1}$) is given by survival/growth transitions from size $x$ and ant state $a$ in year $t$, plus germination out of the seed banks:
\begin{linenomath*}
	$$
	n(x',a')_{t+1} = (\gamma_1 B^1_{t} + \gamma_2 B^2_{t} ) \eta(x') \omega \rho_{0}(a')  + \\
	$$
	$$
	\sum_{a=1}^{A} \int_L^U S(x,a;\pmb{\tau^S_{a}}) G(x',x,a;\pmb{\tau^G_{a}}) \rho(x,a,a';\pmb{\tau^{\epsilon}}) n(x,a)_t dx \\
	$$
\end{linenomath*}

\noindent The first term estimates the number of individuals recruiting from a one or two-year seed bank to a plant of size $x'$ and ant state $a'$ based on the recruit size distribution $\eta(x')$ and the probability of seedling survival ($\omega$) from germination (late summer) to the census (May).
This term is multiplied by $\rho_{0}(a')$, which gives the probability that a new recruit has ant state $a'$ at its first appearance in our census ($\sum_{a'}^{A}\rho_{0}(a')=1$). 
The second term represents all possible transitions from size $x$ and ant $a$ to size $x'$ and ant $a'$, conditioned on survival. 
Survival from initial size $x$ ($S(x,a;\pmb{\tau^S_{a}})$) and growth from size $x$ to $x'$ ($G(x',x,a;\pmb{\tau^G_{a}})$) are both dependent on initial size and ant state. 
As above, these functions include inter-annual variability through year-specific deviates that can vary by ant state ($\pmb{\tau_{a}}$). 
Ant transition function $\rho(a',a,x;\pmb{\tau^{\rho}})$ gives the probability that an individual transitions from ant state $a$ to $a'$ in the next census, conditional on initial size $x$. 
This function includes inter-annual variability through year-specific intercepts which are consistent across ant states ($\pmb{\tau^\rho}$).

\subsection*{Statistical modeling and parameter estimation}
We parameterized the IPM using a series of generalized linear mixed models (GLMMs) in a hierarchical Bayesian framework to serve as vital rate sub-models. 
Vital rate models included spatial and temporal random effects associated with plot and year variation, respectively, and included plant size ($log(cm^3)$; $x,x'$), ant partner state ($a,a'$), or both as fixed-effect predictor variables. 
In addition to vital rate models describing plant demographic performance, we also fit a sub-model to predict transition between ant states conditional on plant size and previous ant state. 
All models were fit using R version 4.2.2 and Rstan package (\cite{Rstancite, Rcite}).
Unless otherwise mentioned, all used vague priors.

The vital rate models take the form of the function:
$$ f(\mu) = \beta_0 + \beta_1 x + ... + u + w $$
\noindent where $u \sim N(0,\sigma_{yr }^{2})$ is the year random effect with year specific deviates of $\sigma_{yr}$ (which parameterizes the $\tau$ vector), and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$. 
Some models deviate from this general model, which will be specified below.

\paragraph{Growth}
The growth sub-model ($G(x',x,a;\pmb{\tau^G_{a}})$) gives the probability of future size given the fixed effects of previous size $x$ and previous ant partner $a$ and random effects of plot $w$ and year $u$ which interacts with ant state $a$. 
We fit this model to the size in year $t+1$ ($y^G$) using the location-scale parameterization of the student $t$ distribution and the identity link function, because in preliminary analyses we found that size transition data were more fat-tailed than a Gaussian distribution could accommodate. 
$$y^G \sim Student T(\hat{\nu},\hat{G},\hat{\sigma}) $$
Student $t$ distributions are dependent on three parameters, $\hat{\nu}$,  $\hat{G}$, and $\hat{\sigma}$.
$\hat{G}$ is the mean of the distribution, and in our case is expressed as a second-order polynomial with ant-size interactions because  preliminary analysis found this was a better fit to slightly concave data. 
$\hat{\nu}$ is the shape parameter which determines how thin the tails of the distribution are.
$\hat{\sigma}$ is the scale parameter which determines how wide the distribution is. 
Both $\hat{\nu}$ and $\hat{\nu}$ are size dependent because preliminary analyses revealed that there was size dependence in both variance and kurtosis.
%$$\hat{G} = \beta_{0}^{G} + \beta_{1}^{G}  x  + \beta_{2}^{G}  a + \beta_{3}^{G}  x  a  + \beta_{4}^{G}  x^2 + \beta_{5}^{G}  a  x^2 + u + w $$
$$\hat{\sigma} = \beta_{0}^{\sigma} + \beta_{1}^{\sigma}  x $$
$$\hat{\nu} = \beta_{0}^{\nu} + \beta_{1}^{\nu}  x $$

\paragraph{Survival}
The survival model ($S(a,x;\pmb{\tau_{a}^{S}})$) estimates the probability of survival from year $t$ to year $t+1$, with fixed effects of the previous size of the cholla $x$ and ant partner $a$ in year $t$ and random effects of plot $w$ and year $u$ which interacts with ant state $a$.
We fit this model to the survival data $y^S$  using a Bernoulli distribution and the logit link function
$y^S \sim Bern(\hat{S})$.
%$$logit(\hat{S}) = \beta_{0}^{S} + \beta_{1}^{S}  x + \beta_{2}^{S}  a + \beta_{3}^{S}  x  a + u + w$$

\paragraph{Reproduction}
The reproduction model ($P(x';\pmb{\tau^{P}})$) estimates the probability of reproducing in year $t+1$, with fixed effects for the size $x'$ in year $t+1$ and random effects of plot $w$ and year $u$.
We fit this model to the reproductive data $y^R$ using a Bernoulli distribution and a logit link function:
$$y^P \sim Bern(\hat{P})$$
%$$logit(\hat{P} = \beta_{0}^{P} + \beta_{1}^{P} \times x' + u + w)$$

\paragraph{No. Flowers}
The total flowers model ($F(x';\pmb{\tau^{F}})$) estimates the total flowers produced by a plant in year $t+1$, with fixed effects of size $x'$ in year $t+1$ and random effects of plot $w$ and year $u$. 
We fit this model to flowerbud count data $y^F$ using a zero-truncated negative binomial distribution with a log transformation:
$$y^{F} \sim 0 Truncated Negative Binom(\hat{F},\hat{\phi})$$
%$$log(\hat{F}) = \beta_{0}^{F} + \beta_{1}^{F} \times x' + u + w$$
$$log(\hat{\phi}) = \beta_{0}^{\phi}$$

\paragraph{Flowerbud viability}
The viability model ($V(a;\pmb{\tau^{V}_{a}})$) estimates theproportion of flowers produced by a plant which are viable (not aborted) in year $t+1$, with fixed effects of ant partner $a$ in year $t$ and random effects of plot $w$ and year $u$ which interacts with ant state $a$.
We fit this model to floral abortion data $y^V$ using a Binomial distribution and a logit link function:
$$y^{V} \sim Binom(\hat{V})$$
%$$logit(\hat{V}) = \beta_{0}^{V} \times a + u + w$$

\paragraph{Ant Transitions}
The ant transition model ($\epsilon(x,a,a';\pmb{\tau^{\epsilon}})$) estimates the probability of a cactus being visited by an ant partner $a'$, with fixed effects of the previous size $x$  and the previous ant partner $a$  in year $t$ and random effects of plot $w$ and year $u$.
We fit this model to ant partner data from year $t+1$ $a'$ using a multinomial distribution with a logit link function: 
$$y^{\epsilon} ~ Multinomial(\hat{\epsilon})$$
%$$logit(\epsilon) = \beta_{0}^{\epsilon} + \beta_{1}^{\epsilon} \times x + \beta_{2}^{\epsilon} \times a + u$$

\paragraph{Parameter estimation}
To obtain posterior estimates of the demographic parameters, we fit models using Markov chain Monte Carlo (MCMC) simulations via STAN run through version 4.0.2 of R \cite{Rcite,Rstancite}. 
For each model, we obtained three chains of 10,000 iterations, each with randomly chosen initial conditions. 
The first 1,500 iterations were discarded. 
We did not thin the chains, thus all samples were retained following burn-in. 
We assessed parameter convergence between and within chains (Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} b). 
To assess the overall model fit we carried out posterior predictive checks to examine how well the fitted model can generate simulated data similar to the real data (Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} a).
All estimated parameters are available in Table \ref{tab:Params}.

\subsection*{IPM Analysis}
We constructed two stochastic IPMs in order to simulate ant combinations that don’t exist in real life with two different versions of environmental variation.
The first is referred to as the stochastic IPM, in which we include random annual variation in the model and allow each ant partner to respond distinctly to annual variation. 
The second is referred to as the stochastic null IPM, in which we include random annual variation in the model, but force each ant partner to respond in the same way to annual variation. 
By analyzing the stochastic model we can estimate the fitness of the cholla under these simulated combinations.
By comparing the stochastic model to the stochastic null model we can identify what mechanisms may explain differences in cholla fitness with different ant partners. 
In this section I first explain how a simple deterministic IPM is developed before diving into how we created our more complex stochastic and stochastic null models.

Analyzing an IPM requires discretizing the continuous IPM kernel into an approximating matrix. 
Size variable $x$ is discretized into $b$ bins, resulting in a $b \times b$ matrix.
In our model there is additional complexity in the form of transitions between ant partners, leading to a matrix size of $A b \times A b$ where $A$ is the number of unique combinations of $a$ and $a'$ (how many possible ant transitions there are).
Our model also has two additional discrete states (year one and year two seed banks) leading to a final matrix size of $A(b+2) \times A(b+2)$.
We used $b = 200$ bins after stability analysis to determine how many bins were needed to consistently estimate the same lambda.
We extend the integration limits $L$ and $U$ (the lowest and highest size values observed) to avoid unintentional eviction \cite{Williams2012}.

Traditionally, in a bayesian IPM, rather than having one set of parameters to populate an approximating matrix and estimate population fitness ($\lambda$), we use 1,000 draws from the posterior distributions of our bayesian statistical models to populate 1,000 approximating matrix and estimate a 1,000 element distribution for $\lambda$.
After preliminary stability analysis we found that 1,000 random posterior draws leads to the same estimate of $\lambda$ as any greater number of posterior draws.
In a stochastic bayesian IPM, the estimation of each individual $\lambda$ in the distribution comes from a geometric mean of many values. 
Rather than a single approximating matrix for each value of $\lambda$, we take the geometric mean of 500 approximating matricies with year-ant-specific effects, resulting in a stochastic version of populationfitness ($\lambda_{S}$).
These year-ant-specific effects are calculated as random effects for each unique combination of year and ant species included in our data. 
While we do not have data on how the ant species actually fluctuate annually, we are able to measure how the effects of the ant species on plants fluctuate annually.
Each of these 500 approximating matricies have the same draw from the posterior distribution, but are assigned the year-ant-specific effects of a random year of our data (year-ant-specific effects are sampled randomly with replacement).
We chose to use 500 year-specific values, because our $\lambda_{S}$ calculations stablilized after 500 year-specific values. 
To calculate a single $\lambda_{S}$ for these 500 year-ant-specific estimates, we take the geometric mean of the \ali{500 dominant eigenvalues}{OK, I am not positive this is correct, but I have forgotten how to correctly describe it. Can we talk through this sentence?} for a single stochastic $\lambda_{S}$.
This process is repeated for each of the 1,000 random posterior draws, leading to a distribution of 1,000 stochastic $\lambda_{S}$ values. 
To calculate the stochastic null bayesian IPM we use the exact same process, but the year-specific values used to calculate the stochastic null $\lambda_{SN}$ are not ant specific, forcing all ant species to have the same effects on their cactus partners in response to environmental variation.

@@

>>>>

Multiple insect herbivores and seed predators specialize on tree cholla \citep{Mann1969}. 
The Cerambycid beetle \textit{Moneilema appressum} and a weevil (Coleoptera: Curculionidae) of the genus \textit{Gerstaekeria} feed on vegetative and reproductive structures as adults and their larvae feed internally. 
A cactus bug, \textit{Narnia pallidicornis} (Hemiptera: Coreidae), feeds on all cholla parts with a preference for the reproductive structures \citep{Miller2006}.
A seed predator, \textit{Cahela ponderosella} (Lepidoptera: Pyralidae), oviposits in open flowers and larvae eat seeds in developing fruits. 
These predators can have significant negative impacts on plant fitness of and depress population growth \citep{Miller2009}.
Prior experiments showed that ant-tended tree cholla experience less herbivory and seed predation than plants from which ants were excluded \citep{Miller2007,Ohm2014}.

\subsection*{Data Collection}
This study is based on long-term demographic data spanning 2004 to 2023 at SEV-LTER. 
From 2004 to 2008, we censused 134 plants distributed across three spatial blocks. 
This initial census group was discontinued in 2009, when we established six 30 $\times$ 30-meter plots and tagged all tree cholla within those plots. 
Two additional 30 $\times$ 30-meter plots were added in 2011, and this group of eight plots has since been censused annually through 2023 (with the exception of 2020 due to the pandemic shutdown). 
For all plants, in May or early June of each year we recorded plant survival since the last survey and, for survivors, we recorded height (cm), maximum crown width (cm), and crown width perpendicular to the maximum (cm).
Size measurements were used to calculate plant volume ($cm^3$) based on the volume of an elliptical cone. 
We measured reproduction as counts of viable and aborted flowerbuds. 
We recorded the ant species present (or vacancy if no ants present).
Occurrences of more than one ant species on one plant were rare (less than 5\% of observations), and for the purpose of this analysis we classified the plant as being occupied by the more abundant species. 
Plots were searched for new recruits each year, and these were added to the census.
In total, the data set included 1141 unique individuals and \tom{19 plant-year observations}{It is 19 years but the number of plant-year observations should be the total number of rows in the data set}. 
%These data were used to fit vital rate models (survival, growth, reproduction) as functions of plant size and ant occupancy state.

We used additional, smaller data sets from previously published studies to estimate seed and seed bank parameters. 
Ohm et al. \citeyear{Ohm2014} provide data on the number of seeds per fruit for plants tended by \textit{L. apiculatum}, \textit{C. opuntiae}, or no ants (experimental exclusion), accounting for their effects on pollinator visitation. 
Elderd and Miller \citeyear{elderd2016quantifying} provide data on seed entry to the seed bank and seedling germination and survival rates.

\subsection*{Multi-state Integral Projection Model}
Integral Projection Models describe population dynamics in discrete time, with functions that relate vital rates to continuous state variables \citep{ellnerbook}. 
While IPMs are a natural choice for populations with continuous size structure, they can also be modified to accommodate a combination of continuous and discrete state variables, as we do here. 
We constructed a stochastic, multi-state IPM that stitches together population structure associated with plant size and ant state, allowing us to determine the individual fitness effects of each ant species and the composite effects of multiple partners, with ant transition dynamics and inter-annual variability modeled explictly.

Given the low frequency of ant occupancy states other than \textit{L. apiculatum} and \textit{C. opuntiae} (\textless8\% of observation) we combined all other ants into an ``other'' category, such that our multi-state IPM included four ant states: vacant, \textit{L. apiculatum}, \textit{C. opuntiae}, and other. 
The ``other'' category was made of \textit{Forelius pruinosus} (3.5\% of observations), unidentified species belonging to the genera \textit{Camponotus} (0.9\%), \textit{Aphaenogaster} (0.4\%), \textit{Myrmecocystus} (0.08\%), \textit{Tetramorium} (0.02\%), \textit{Brachymyrmex} (0.02\%), and additional ants not identified to genus or species (2.8\%).

Ant state is included as a predictor variable in IPM sub-models where there are biologically realistic pathways through which ants could impact the outcome of that process. 
For example, ant partners defend cacti from herbivores, and prior experimental work indicates that ant tending can reduce vegetative tissue loss and floral abortion.
Therefore, ant state was included in sub-models for survival, growth, and flowerbud viability. 
In contrast, we have no reason to expect that ant tending can directly influence the probability of flowering and flowerbud production, independently of its influence on plant size, so these sub-models included plant size but not ant state as predictor variables.

We modeled the tree cholla life cycle using continuously size-structured plants where $n(x,a)_{t}$ gives the number of plants of size $x$  and ant state $a$ in year $t$, plus two discrete seed banks ($B^1_{t}$ and $B^2_{t}$) corresponding to 1 and 2-year old seeds. 
Seed bank dynamics are given by:

\begin{linenomath*}
	$$
	B^1_{t+1} = \delta \sum_{a=1}^{A} \int_L^U  \kappa(a') P(x;\pmb{\tau}^P) F(x;\pmb{\tau}^F) V(a;\pmb{\tau}^V_{a}) n(x,a)_{t} dx \\
	$$
	$$
	B^2_{t+1} =  (1 - \gamma_1)B^1_{t}\\
	$$
	\label{eqn:IPM1}
\end{linenomath*}

\noindent %In these equations $x$ and $x'$ indicate the size of a plant in years $t$ and $t+1$ respectively, $a$ and $a'$ indicate the ant partner of a plant in years $t$ and $t+1$.
Functions $P(x;\pmb{\tau}^P)$ and $F(x;\pmb{\tau}^F)$ give the probability of flowering in year $t$ and the number of flowerbuds produced in year $t$, respectively, by plants of size $x$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau}^V_{a})$) and the number of seeds per fruit \tom{($\kappa(a)$)}{This is correct but in the equation above it is shown as $\kappa(a')$ -- I am flagging that so you can update the equation and check that the code is consistent.} is dependent on ant state $a$. 
The vectors $\pmb{\tau}$ give year-specific deviates (mean zero) and appear in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and subscripts indicate that deviates are specific to plants in ant state $a$.
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible ant states ($A=4$) giving total seed production. 
Seeds are multiplied by the probability of seed dispersal and survival ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

For the above-ground part of the life cycle, the number of plants of size $x'$ and ant state $a'$ in year $t+1$ ($n(x',a')_{t+1}$) is given by survival/growth transitions from size $x$ and ant state $a$ in year $t$, plus germination out of the seed banks:
\begin{linenomath*}
	$$
	n(x',a')_{t+1} = (\gamma_1 B^1_{t} + \gamma_2 B^2_{t} ) \eta(x') \omega \rho_{0}(a')  + \\
	$$
	$$
	\sum_{a=1}^{A} \int_L^U S(x,a;\pmb{\tau^S_{a}}) G(x',x,a;\pmb{\tau^G_{a}}) \rho(x,a,a';\pmb{\tau^{\epsilon}}) n(x,a)_t dx \\
	$$
	\label{eqn:IPM2}
\end{linenomath*}

\noindent The first term in Eq. \tom{\ref{eqn:IPM2}}{We should label equations. I am not sure why the equation label is not working here and I did not try to figure it out. It is probably something with the linenomath formatting.} estimates the number of individuals recruiting from a one or two-year seed bank to a plant of size $x'$ and ant state $a'$ based on the recruit size distribution $\eta(x')$ and the probability of seedling survival ($\omega$) from germination (late summer) to the census (May).
This term is multiplied by $\rho_{0}(a')$, which gives the probability that a new recruit has ant state $a'$ at its first appearance in our census ($\sum\rho_{0}(a')=1$). 
The second term represents all possible transitions from size $x$ and ant $a$ to size $x'$ and ant $a'$, conditioned on survival. 
Survival from initial size $x$ ($S(x,a;\pmb{\tau}^S_{a})$) and growth from size $x$ to $x'$ ($G(x',x,a;\pmb{\tau}^G_{a})$) are both dependent on initial size and ant state. 
As above, these functions include inter-annual variability through year-specific deviates that can vary by ant state ($\pmb{\tau}_{a}$). 
Ant transition function $\rho(a',a,x;\pmb{\tau}^{\rho})$ gives the probability that an individual transitions from ant state $a$ to $a'$ in the next census, conditional on initial size $x$. 
This function includes inter-annual variability through year-specific intercepts which are consistent across ant states ($\pmb{\tau}^\rho$).

\subsection*{Statistical modeling and parameter estimation}
We parameterized the IPM using a series of generalized linear mixed models in a hierarchical Bayesian framework to serve as vital rate sub-models. 
Vital rate models included spatial and temporal random effects associated with plot and year variation, respectively, and included plant size ($log(cm^3)$; $x,x'$), ant partner state ($a,a'$), or both as fixed-effect predictor variables. 
In addition to vital rate models describing plant demographic performance, we also fit a sub-model to predict transitions between ant states conditional on plant size and previous ant state. 
All models were fit using R version 4.2.2 and Rstan package (\cite{Rstancite, Rcite}).

Vital rate models for survival, growth, probability of flowering,  take the form of the function:
$$ f(\mu) = \beta_0 + \beta_1 x + ... + u + w $$
\noindent where $u \sim N(0,\sigma_{yr }^{2})$ is the year random effect with year specific deviates of $\sigma_{yr}$ (which parameterizes the $\tau$ vector), and $w \sim N(0,\sigma_{plot}^{2})$ is the plot random effect with plot specific variance of $\sigma_{plot}$. 
Some models deviate from this general model, which will be specified below.

\paragraph{Growth}
The growth sub-model ($G(x',x,a;\pmb{\tau^G_{a}})$) gives the probability of future size given the fixed effects of previous size $x$ and previous ant partner $a$. 
We fit this model to the size in year $t+1$ ($y^G$) using the skewed normal distribution to account left-skewed size transitions:
$$y^G \sim Skewed Normal(\hat{G},\sigma,\alpha) $$
$$\hat{G} = \beta^G_0 + \beta^G_1 x + \beta^G_2 a + \beta^G_3 ax + \beta^G_4 x^2 + \beta^G_5 x^2a + u^G_{a} + w^G $$
The \tom{mean}{I rewrote this for the skewed normal but I cannot remember if the location parameter is the mean, so you should double check this.} $\hat{G}$ is a second-order polynomial with ant-size interactions because  preliminary analysis found this was an improvement over a linear relationship. 
$u^G_{a} \sim N(0,\sigma_{year}^{2})$ is the year random effect for each ant state (which parameterizes the $\pmb{\tau}^G_{a}$ vector), and $w^G \sim N(0,\sigma_{plot}^{2})$ is the plot random effect. 
Parameters $\hat{\sigma}$ and $\hat{\alpha}$  control residual variance and skewness, respectively, and were defined as linear functions of initial size $x$.

\paragraph{Survival}
The survival sub-model ($S(a,x;\pmb{\tau}_{a}^{S})$) estimates the probability of survival from year $t$ to year $t+1$, with fixed effects of size $x$ and ant partner $a$ in year $t$.
We fit this model to the survival data $y^S$  using a Bernoulli distribution and the logit link function, with year and plot random effects as above. 
$$y^S \sim Bern(\hat{S})$$
$$logit(\hat{S}) = \beta_{0}^{S} + \beta_{1}^{S}  x + \beta_{2}^{S}  a + \beta_{3}^{S}  x  a + u^S_{a} + w^S$$

\paragraph{Reproduction}
The flowering sub-model ($P(x';\pmb{\tau}^{P})$) estimates the probability of reproducing in year $t+1$, with fixed effects for the size $x'$ in year $t+1$ and random effects of plot and year.
We fit this model to the reproductive status data (vegetative or flowering) using a Bernoulli distribution and a logit link function, similar to the survival model above but with no ant effects. 
The flower bud function $F(x';\pmb{\tau}^{F})$ estimates the total flowers produced by a reproducing plant in year $t+1$, with fixed effects of size $x'$ in year $t+1$. 
We fit this model to flowerbud count data $y^F$ (sum of viable and aborted) using a zero-truncated negative binomial distribution with a log link and normally distributed year and plot random effects as above:
$$y^{F} \sim NegBinom(\hat{F},\hat{\phi})$$
$$log(\hat{F}) = \beta_{0}^{F} + \beta_{1}^{F} \times x' + u^F + w^F$$

The flowerbud viability sub-model ($V(a;\pmb{\tau}^{V}_{a})$) estimates the \tom{proportion of flowers produced by a plant that are viable (not aborted) in year $t+1$, with fixed effects of ant partner $a$ in year $t$}{The way this is written here is not correct. Viability in year t should be a function of ants in year t. I would double check that this is what the model is actually doing.}.
We fit this model to \tom{floral abortion data $y^V$}{The way this is written it's unclear whether you are modeling probability of viability or probability of abortion} using a binomial distribution with $N$ trials (total number of flower buds) and a logit link function:
$$y^{V} \sim Binom(\hat{V},N)$$
$$logit(\hat{V}) = \beta_{0}^{V} \times a + u + w$$

Additional reproductive parameters for the number of seeds per fruit, probability of entry to the seed bank, germination rates, and recruit size were estimated following methods described in Appendix XX.

\paragraph{Ant Transitions}
The ant transition model ($\epsilon(x,a,a';\pmb{\tau}^{\epsilon})$) estimates the probability of a cactus being occupied by ant partner $a'$ in year $t+1$, with fixed effects of the previous size $x$  and the previous ant partner $a$  in year $t$ and random effects of plot $w$ and year $u$.
We fit this model to ant partner data $y^{\epsilon}$ using a multinomial distribution with a logit link function: 
$$y^{\epsilon} \sim Multinomial(\hat{\epsilon})$$
$$logit(\epsilon) = \beta_{0}^{\epsilon} + \beta_{1}^{\epsilon} \times x + \beta_{2}^{\epsilon} \times a + u$$

\paragraph{Parameter estimation}
We fit models using Markov chain Monte Carlo (MCMC) simulations via STAN run through version 4.0.2 of R \cite{Rcite,Rstancite}. 
We used vague priors for all parameters. 
For each model, we obtained three chains of 10,000 iterations, discarding the first 1,500 iterations. 
%We did not thin the chains, thus all samples were retained following burn-in. 
We visually assessed parameter convergence between and within chains (Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} b) and assessed overall model fit with posterior predictive checks to examine how well the fitted model can generate simulated data similar to the real data (Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} a).
%All estimated parameters are available in Table \ref{tab:Params}.

\subsection*{IPM Analysis}
Analyzing an IPM requires discretizing the continuous IPM kernel into an approximating matrix. 
Size variable $x$ is discretized into $b$ bins, resulting in a $b \times b$ matrix.
In our model there is additional complexity in the form of transitions between ant partners, leading to a matrix size of $A b \times A b$.
Our model also has two additional discrete states (year one and year two seed banks) leading to a final matrix size of $A(b+2) \times A(b+2)$.
We used $b = 200$ bins after stability analysis to determine how many bins were needed to consistently estimate the same lambda.
We extended the integration limits beyond the minimum ($L$) and maximum ($U$) observed sizes to avoid unintentional eviction \cite{Williams2012}.

For stochastic analyses, we estimated the approximating matrix corresponding to each $t$ to $t+1$ transition year. 
To estimated population mean fitness in a stochastic environment ($\lambda_{S}$) we simulated population dynamics for 500 years by randomly sampling among the \tom{19}{Is this right? The number of transition matrices should be one fewer than then number of years, but I don't remember what happens with 2020.} annual transition matrices, discarding the first 100 years to minimize the influence of initial conditions. 
Sampling observed transition matrices (rather than independently sampling regression coefficients) produces models that realistically capture inter-annual variation by preserving correlations between vital rates \cite{metcalf2015statistical}.
We tallied the total population size at each time step as  $N_{t} = B^1_{t} + B^2_{t} + \sum_{a=1}^{A}\int n(x,a)_{t}dx$ and calculated the stochastic growth rate as $log(\lambda_S) = E[log(\frac{N_{t}}{N_{t+1}})]$ \cite{rees2009integral}.
We propagated uncertainty from the vital rate models using 1000 draws from the joint posterior distribution of model parameters, resulting in a posterior distribution of $\lambda_{S}$ and other derived quantities.

Using the fully parameterized multi-state IPM, we conducted simulation ``experiments'' to test how diversity and identity of ant partners influenced plant fitness ($\lambda_{S}$). 
From the fullest version of the model corresponding to the observed assemblage of partners, we created model versions

@@

We constructed two versions of the stochastic, multi-state IPMs that allowed us to explore two different scenarios of environmental variation. 
Under the portfolio effect hypothesis, partner diversity may confer a fitness advantage when the benefits of alternative partners are not perfectly synchronized across temporal environmental variation.

The first is referred to as the stochastic IPM, in which we include random annual variation in the model and allow each ant partner to respond distinctly to annual variation. 
The second is referred to as the stochastic null IPM, in which we include random annual variation in the model, but force each ant partner to respond in the same way to annual variation. 
By analyzing the stochastic model we can estimate the fitness of the cholla under these simulated combinations.
By comparing the stochastic model to the stochastic null model we can identify what mechanisms may explain differences in cholla fitness with different ant partners. 
In this section I first explain how a simple deterministic IPM is developed before diving into how we created our more complex stochastic and stochastic null models.
